---
title: "Lab 12 - Correlation and Visualization"
author: "Ajul Thomas"
date: "2023-10-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Exercises
Exercise 1

1)Read the data from the “advertising.csv” file and get yourself familiar with the included variables.

```{r}

adv_df <- read_csv("data/Advertising.csv")

adv_df

```


    
2)Remove the un-named data variable(s), if any

```{r}

adv_df <- adv_df[-1] 

adv_df


```


3)Check the statistical summary of the “sales” variable (hint: you can use the summary function)

```{r}

summary(adv_df)


```


4)How about using the summarise() function?

```{r}

adv_df %>% 
  group_by() %>% 
  summarise()


```



Exercise 2

Based on the previous exercise, complete the following tasks:

1)Plot scatter plots between the sales and each of the other variables in the data.

```{r}

ggplot(adv_df, aes(x=sales)) +
  geom_point(aes(y=newspaper), color="red")
  
ggplot(adv_df, aes(x=sales)) +
  geom_point(aes(y=radio), color="blue")

ggplot(adv_df, aes(x=sales)) +
  geom_point(aes(y=TV), color="purple")

```


2)Add a regression line using stat_smooth(method = lm) to the previous plots? What does lm provide?

```{r}

ggplot(adv_df, aes(x=sales, y=newspaper)) +
  geom_point( color="red") +
  geom_smooth(method=lm)
  
ggplot(adv_df, aes(x=sales, y=radio)) +
  geom_point(color="blue") +
  geom_smooth(method=lm)

ggplot(adv_df, aes(x=sales, y=TV)) +
  geom_point(color="purple") +
  geom_smooth(method=lm)


```

3)Compute the correlation between sales variable and each of the other variables independently, using cor(var1, var2) function.

```{r}

r_tv <- cor(adv_df$sales, adv_df$TV)
r_tv

r_news <- cor(adv_df$sales, adv_df$newspaper )
r_news

r_radio <- cor(adv_df$sales, adv_df$radio)
r_radio

```

4)Generate the correlation matrix for all the variables together.


```{r}

cor(adv_df)

```

5) Visualise the correlation matrix using ggcorr(…) function, by replacing … with the data frame.

```{r}


install.packages("GGally")
library(GGally)

ggcorr(adv_df, label = T, label_alpha = T )

```


Exercise 3

We can also use the correlation matrix to test the linear relationship between variables, using cor.test(var1, var2, method), where method can be “Pearson”, “Spearman”, etc.

1)Use the cor.test function to test the estimated correlation between sales and TV variables.

```{r}

cor_test_results <- cor.test(adv_df$sales, adv_df$TV, method="pearson")
cor_test_results

```

2)How would you interpret the p-value and estimate values?

```{r}



```


3)Can you extract the p-value and estimate coefficients from the output of the correlation test?

```{r}

# p value
cor_test_results$p.value

# correlation coefficient
cor_test_results$estimate

```


Exercise 4 - Generating code for a given plot

1)Please download “weather.csv” data from the contents of week 11’)’s module.


```{r}

weather_df <- read_csv('data/weather_data.csv')

weather_df

str(weather_df)

```


2)Given these data, can you write the code to generate the following plot? What are the steps would you follow?

```{r}

df_to_plot <- weather_df %>% 
  # separate(col = Date, into = c('Year', 'Month'), sep = '\\-', remove = F) %>% 
  mutate(avg_temp=(`9am_Temperature` + `3pm_Temperature`)/2) %>% 
  group_by(month) %>% 
  summarise(
    min_temp = min(Minimum_temperature, na.rm = T),
    max_temp = max(Maximum_temperature, na.rm = T),
    mean_temp = mean(avg_temp, na.rm = T)
  )

months = c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "July", "Aug", "Sep", "Oct", "Nov", "Dec")


df_to_plot_2 <- df_to_plot %>% 
  gather(key="metric", value="value", -month)

ggplot(df_to_plot_2, aes(x=month, y=value, group=metric, colour=metric)) +
         scale_x_discrete(limits=months) + 
         geom_point() +
         geom_line() +
         labs(title = "The average, minimum and maximum Temperature in 2019", x = "Months", y = "Value (in degree celsius)")

```


