---
title: 'Assignment 1 - Solution'
author name: 'Ajul Thomas'
author ID: 'u3253992'
output:
  html_document: rmdformats::readthedown
editor_options: 
  markdown: 
    wrap: 100
---
***

## Data Description
The following code generates the required dataset for this assignment. It will create a data frame with 92 rows and 13 columns. The variables of this data frame are as following:

- `Date`: Date of the reading (in YYYY-MM-DD format)
- `Temperature`: Temperature reading in Celsius
- `Humidity`: Humidity reading as a percentage
- `Pressure`: Atmospheric pressure in millibars
- `WindSpeed`: Wind speed in kilometers per hour
- `WindDirection`: Wind direction
- `DewPoint`: Dew point temperature in Celsius
- `CloudCover`: Cloud cover as a percentage
- `Precipitation`: Precipitation amount in millimeters
- `Visibility`: Visibility distance in kilometers
- `UVIndex`: UV index reading
- `condition`: The global weather condition over the day
- `Location`: The city of the recorded data

```{r, eval=FALSE}
# Set the seed for reproducibility
set.seed(123)

# Create a sequence of dates for the month of March 2023
dates <- seq(from = as.Date("2023-03-01"), to = as.Date("2023-05-31"), by = 1)

# Create a data frame to store the weather data
weather <- data.frame(
  Date = dates,
  Temperature = round(runif(length(dates), 15, 25), 1),
  Humidity = sample(50:100, length(dates), replace = TRUE),
  Pressure = sample(995:1015, length(dates), replace = TRUE),
  WindSpeed = sample(5:20, length(dates), replace = TRUE),
  WindDirection = sample(c("N", "NE", "E", "SE", "S", "SW", "W", "NW"), length(dates), replace = TRUE),
  DewPoint = round(runif(length(dates), 10, 15), 1),
  CloudCover = sample(0:100, length(dates), replace = TRUE),
  Precipitation = round(runif(length(dates), 0, 15), 1),
  Visibility = sample(5:20, length(dates), replace = TRUE),
  UVIndex = sample(1:5, length(dates), replace = TRUE),
  Condition = sample(c("Sunny", "Partly Cloudy", "Rainy", "Snowy"), length(dates), replace = TRUE),
  Location = rep(c("Sydney", "Canberra"), each = length(dates)/2)
)
# do the required here to check the generated data

# check contenets of weather dataframe
print(weather)

# to check dimensions of dataframe
dim(weather)

# check structure of data frame
str(weather)


```

***

## Submission for Part A: Data Understanding
Please follow this structure:

```{asis, echo=TRUE}

----------------------------------

1- The data set description goes here

This dataset contains weather-related information for the cities of Sydney and Canberra for the month of March, April and May in 2023. It comprises 92 observations and includes 13 variables. The variables 'WindDirection', 'Condition' and 'Location' are categorical and the rest are quantitative.

This dataset can be used for various types of weather analysis specific to the cities of Sydney and Canberra during the month of March. Researchers and analysts can examine temperature variations, humidity levels, precipitation patterns, and wind characteristics to understand the unique weather conditions in these two cities during that time frame. It can also be valuable for assessing how different weather conditions may have impacted daily life, outdoor activities, or local businesses in Sydney and Canberra during March. Furthermore, the data can be used to build predictive models for short-term weather forecasting in these specific locations.

----------------------------------
```

```{r}
#----------------------------------#

# 2-  The code for task 2 goes here

summary(weather)


#----------------------------------#

# 3-  The code for task 3 goes here

hist(weather$Temperature)

#----------------------------------#
```
```{asis, echo=TRUE}
4- The reflection and notes for task 2 and 3 goes here

For task 2, the 'summary()' function is really useful in getting an overall overview of the nature of various variable within the dataset we are dealing with. For instance, if we are looking at the Temperature variable, we can observe that the maximum temperature observed is 24.90 degree celcius and the min tempurature recorded is 15 degree celcius, with the mean at 20.05 degree celcius. Also, it is also extremely useful in understanding the categorical variables in the dataset. we have three categorical variables in weather dataset, 'WindDirection', 'Condition' and 'Location'.


For task 3, the histogram of the temperature reading suggests that the distribution of the data is bimodal with one mode between 19 degree celcius and 20 degree celcius and another mode between 23 degree celcius and 24 degree celcius.


----------------------------------

```

## Submission for part B: Vector and Matrix Manipulation
Please follow this structure:

```{r}

library(tidyverse)

# to check the data type of the data column
class(weather$Date)

# Extract the month and year from the "Date" column
weather$YearMonth <- format(weather$Date, "%Y-%m")

# extract year, month and date
weather <- weather %>%
  separate(col=Date, into=c("Year","Month","Day"), sep="\\-", remove=FALSE)

#----------------------------------#
# 1-  The code for task 1 goes here
# Create a vector containing the average temperature readings for each month.

# using group_by and summarize to find the monthly average temperature
monthly_average_temperature <- 
  (
  weather %>%
    group_by(Month) %>%
    summarize(average_temp = mean(Temperature))
  )$average_temp

cat("Q1: Monthly average temperatures: ")
print(monthly_average_temperature)



#----------------------------------#

# 2-  The code for task 2 goes here
# Create two vector containing the average humidity readings for each month; one for Sydney and another for Canberra.

canberra_monthly_avg_humidity <- 
  (
    weather %>%
      filter(Location=='Canberra') %>%
      group_by(Month) %>%
      summarise(average_humidity=mean(Humidity))
  )$average_humidity

sydney_monthly_average_humidity <- 
  (
    weather %>%
      filter(Location=='Sydney') %>%
      group_by(Month) %>%
      summarise(average_humidity=mean(Humidity))
  )$average_humidity

cat("Q2: Monthly average humidity for Sydney and Canberra: \n")
print(canberra_monthly_avg_humidity)
print(sydney_monthly_average_humidity)

#----------------------------------#

# 3-  The code for task 3 goes here

# Create a matrix containing the average temperature, humidity, pressure, and wind speed readings for each month.


# first let's calculate the average values of these parameters using tidyverse functions group_by and summarize.
monthly_averages_df <- weather %>%
  group_by(Month) %>%
  summarize(
    average_temp = mean(Temperature), 
    average_humidity = mean(Humidity),
    average_pressure = mean(Pressure),
    average_wind_speed = mean(WindSpeed)
    )

monthly_averages_df <- monthly_averages_df %>% select(-Month)

# as the next step, we will convert the df into a matrix
# reference: https://stat.ethz.ch/R-manual/R-devel/library/base/html/data.matrix.html
monthly_averages <- data.matrix(monthly_averages_df)

# is.matrix(monthly_averages)
print("Matrix containing the average temperature, humidity, pressure, and wind speed readings for each month: ")
monthly_averages




#----------------------------------#

# 4-  The code for task 4 goes here
# Create another matrix containing the average temperature, humidity, pressure, and wind speed readings for each city.


city_averages_df <- weather %>%
  group_by(Location) %>%
  summarize(
    average_temp = mean(Temperature), 
    average_humidity = mean(Humidity),
    average_pressure = mean(Pressure),
    average_wind_speed = mean(WindSpeed)
    )
city_averages_df <- city_averages_df %>% select(-Location)

# as the next step, we will convert the df into a matrix
# reference: https://stat.ethz.ch/R-manual/R-devel/library/base/html/data.matrix.html
city_averages <- data.matrix(city_averages_df)

# is.matrix(city_averages)
print("Matrix containing the average temperature, humidity, pressure, and wind speed readings for each city: ")
city_averages



#----------------------------------#

# 5-  The code for task 5 goes here
# Create an array containing the average temperature, humidity, pressure, wind speed, and UV index readings for each day of each month.




#----------------------------------#

# 6-  The code for task 6 goes here
# Create an array containing the average temperature, humidity, pressure, wind speed, and UV index readings for each day of each month, for each city.




#----------------------------------#

# 7-  The code for task 7 goes here

# Use matrix multiplication to calculate the product of the transpose of the matrix in task 3 with the vector in task 1.

# vector from task 1
cat("Task 1: vector containing monthly average temperatures: \n")
cat("------------------------------------------------------\n")
print(monthly_average_temperature)
cat("\n")

# matrix from task 3
cat("Task 3: matrix containing the average temperature, humidity, pressure, and wind speed readings for each month: \n")
cat("-------------------------------------------------------------------\n")
print(monthly_averages)
cat("\n")

# matrix transpose
cat("Transpose of the matrix : \n")
cat("------------------------------------------------------\n")
mat_transpose <- t(monthly_averages)
print(mat_transpose)
cat("\n")

# matrix multiplication
product <- mat_transpose %*% monthly_average_temperature
rownames(product) <- NULL
cat("\n Product of matrix multiplication:\n")
cat("---------------------------------\n")
print(product)

#----------------------------------#

```

*** 

## Submission for part C: Looping and Conditional Statements
Please follow this structure:

```{r}
#----------------------------------#

# 1-  The code for task 1 goes here
# Write a loop that calculates the average pressure reading for each month and stores the results in a vector.

# Initialize lists with month names as names
monthly_pressure_readings_sum <- list()
monthly_pressure_readings_count <- list()

for (i in 1:nrow(weather)) {
  key <- weather$Month[i]
  pressure <- as.numeric(weather$Pressure[i])
  
  # Check if the key (month) is already in the lists, and if not, initialize it
  if (!(key %in% names(monthly_pressure_readings_sum))) {
    monthly_pressure_readings_sum[[key]] <- 0
    monthly_pressure_readings_count[[key]] <- 0
  }
  
  # Update the sum and count for the corresponding month
  monthly_pressure_readings_sum[[key]] <- monthly_pressure_readings_sum[[key]] + pressure
  monthly_pressure_readings_count[[key]] <- monthly_pressure_readings_count[[key]] + 1
}

average_monthly_pressure <- c()

for (month in names(monthly_pressure_readings_sum)) {
  average_value <- monthly_pressure_readings_sum[[month]]/monthly_pressure_readings_count[[month]]
  average_monthly_pressure <- c(average_monthly_pressure,average_value)
}

average_monthly_pressure



#----------------------------------#

# 2-  The code for task 2 goes here

# Use a conditional statement to determine how many days had a temperature above 25 degrees Celsius in Sydney.

# variable to store the no. of instances where the Sydney temp is greater than 25 degree celcius
high_temp_count <- 0

# looping through df
for (i in 1:nrow(weather)) {
  # stores location
  location <- weather$Location[i]
  
  # stores the temperature
  temp <- weather$Temperature[i]
  
  # checks if location is Sydney and temp is greater than 25
  if(location == 'Sydney' & temp > 25) {
    # updates the temp count
    high_temp_count = high_temp_count + 1
  }
}

# print the output
print(high_temp_count)

# to validate
# weather %>%
#   filter(Location == 'Sydney' & Temperature > 25)



#----------------------------------#

# 3-  The code for task 3 goes here
# Use a for loop to calculate the average humidity for the days with a temperature below 21 degrees Celsius in Canberra.

humidity_sum <- 0
days_count <- 0

for (i in 1:nrow(weather)) {
  # stores location
  location <- weather$Location[i]
  
  # stores the temperature
  temp <- weather$Temperature[i]
  
  # checks if location is Sydney and temp is greater than 25
  if(location == 'Canberra' & temp < 21) {
    
    # update days count
    days_count <- days_count + 1
    
    # sum humidity to calculate average
    humidity_sum <- humidity_sum + weather$Humidity[i]
  }
}

average_humidity <- humidity_sum/days_count

average_humidity

# to validate
# weather %>%
#   filter(Location == 'Canberra' & Temperature < 21) %>%
#   summarise(mean_humidity=mean(Humidity))



#----------------------------------#

# 4-  The code for task 4 goes here


# variable to store the high uv days
high_uv_days <- 0

for (i in 1:nrow(weather)) {
  # stores location
  location <- weather$Location[i]
  
  # stores the temperature
  uv_index <- weather$UVIndex[i]
  
  # checks if location is Sydney and temp is greater than 25
  if(location %in% c('Canberra', 'Sydney') & uv_index > 7) {
    
    # update days count
    high_uv_days <- high_uv_days + 1
  }
}

print(high_uv_days)

# to validate
weather %>%
  filter(Location %in% c('Canberra', 'Sydney') & UVIndex > 7) %>%
  summarise(count=n())

#----------------------------------#
```

*** 

## Submission for part D: Data Frame Manipulation
Please follow this structure:

```{r}
#----------------------------------#

# 1-  The code for task 1 goes here
# Load the 5 files using read.csv or read_csv functions and combine them into one data frame. 
# Please note that the first 7 rows of each files need to be ignored while loading the data.

# store filenames
file_names <- c('201812.csv', '201811.csv', '201810.csv', '201809.csv', '201808.csv')

create_file_paths <- function(file_name) {
  file_path <- paste("data",file_name, sep = "/")
  return(file_path)
}

file_paths <- sapply(file_names, create_file_paths)

custom_read_csv <- function(file_path) {
  return(read.csv(file_path, skip=7))
}

data_list <- lapply(file_paths, custom_read_csv)
dim(data_list)

combined_df <- data.frame()

for(df in data_list) {
  print(dim(df))
  combined_df <- rbind(combined_df, df)
}


#----------------------------------#

# 2-  The code for task 2 goes here
# Check the dimensions of the combined data frame.

dim(combined_df)

#----------------------------------#

# 3-  The code for task 3 goes here
# Write a for loop to check the structure 'str()' and summary of each column.

col_names <- colnames(combined_df)

for(col in col_names) {
  cat("Column Name: ", col, "\n")
  cat('-------------------------\n')
  
  cat("Structure: \n")
  cat(str(combined_df[[col]]))
  cat("\n")
  
  cat("Summary: \n")
  print(summary(combined_df[[col]]))
  
  cat("\n")
}

#----------------------------------#

# 4-  The code for task 4 goes here
# Write a code to remove the variables, which have no data at all 
# (i.e. all the records in these variables are NAs).

get_empty_col <- function(col_name) {
  if(all(is.na(combined_df[[col_name]]))) {
    print(col_name)
    return(col_name)
  }
  return(NA)
}

empty_cols <- sapply(col_names, get_empty_col)
empty_cols <- empty_cols[!is.na(empty_cols)]

combined_df <- combined_df %>%
  select(-empty_cols)

#----------------------------------#
# 5-  The code for task 5 goes here
# Write code to change the column names to have no spaces between the words 
# and replace these spaces with underscore the _ character.

colnames(combined_df) <- gsub(" ", "_", colnames(combined_df))

colnames(combined_df)
#----------------------------------#

# 6-  The code for task 6 goes here
# Write code to create two new columns for the month and year from the Date column.

class(combined_df$Date)

combined_df$Date <- as.Date(combined_df$Date, format = "%d/%m/%Y")

combined_df <- combined_df %>%
  separate(col = Date, into = c("Month", "Year"), remove = FALSE)

#----------------------------------#

# 7-  The code for task 7 goes here
# Write code to replace the remaining NAs with the median of the values of each column.


combined_df <- combined_df %>%
  mutate(across(where(is.numeric), ~replace_na(.,median(.,na.rm=TRUE))))

# reference: https://www.statology.org/r-replace-na-with-median/

#----------------------------------#

# 8-  The code for task 8 goes here
# Write code to save the combined weather data file with all of these changes to a single file.

write.csv(combined_df, file = "combined_weather.csv")

#----------------------------------#
```

***

## Overall Reflection

Your overall reflection about this assignment goes here ...
