---
title: 'Assignment 1 - Solution'
author name: 'Ajul Thomas'
author ID: 'u3253992'
output:
  html_document: rmdformats::readthedown
editor_options: 
  markdown: 
    wrap: 100
---
***

## Data Description
The following code generates the required dataset for this assignment. It will create a data frame with 92 rows and 13 columns. The variables of this data frame are as following:

- `Date`: Date of the reading (in YYYY-MM-DD format)
- `Temperature`: Temperature reading in Celsius
- `Humidity`: Humidity reading as a percentage
- `Pressure`: Atmospheric pressure in millibars
- `WindSpeed`: Wind speed in kilometers per hour
- `WindDirection`: Wind direction
- `DewPoint`: Dew point temperature in Celsius
- `CloudCover`: Cloud cover as a percentage
- `Precipitation`: Precipitation amount in millimeters
- `Visibility`: Visibility distance in kilometers
- `UVIndex`: UV index reading
- `condition`: The global weather condition over the day
- `Location`: The city of the recorded data

```{r, eval=FALSE}
# Set the seed for reproducibility
set.seed(123)

# Create a sequence of dates for the month of March 2023
dates <- seq(from = as.Date("2023-03-01"), to = as.Date("2023-05-31"), by = 1)

# Create a data frame to store the weather data
weather <- data.frame(
  Date = dates,
  Temperature = round(runif(length(dates), 15, 25), 1),
  Humidity = sample(50:100, length(dates), replace = TRUE),
  Pressure = sample(995:1015, length(dates), replace = TRUE),
  WindSpeed = sample(5:20, length(dates), replace = TRUE),
  WindDirection = sample(c("N", "NE", "E", "SE", "S", "SW", "W", "NW"), length(dates), replace = TRUE),
  DewPoint = round(runif(length(dates), 10, 15), 1),
  CloudCover = sample(0:100, length(dates), replace = TRUE),
  Precipitation = round(runif(length(dates), 0, 15), 1),
  Visibility = sample(5:20, length(dates), replace = TRUE),
  UVIndex = sample(1:5, length(dates), replace = TRUE),
  Condition = sample(c("Sunny", "Partly Cloudy", "Rainy", "Snowy"), length(dates), replace = TRUE),
  Location = rep(c("Sydney", "Canberra"), each = length(dates)/2)
)
# do the required here to check the generated data

# check contenets of weather dataframe
print(weather)

# to check dimensions of dataframe
dim(weather)

# check structure of data frame
str(weather)


```

***

## Submission for Part A: Data Understanding
Please follow this structure:

```{asis, echo=TRUE}

----------------------------------

1- The data set description goes here

This dataset contains weather-related information for the cities of Sydney and Canberra for the month of March, April and May in 2023. It comprises 92 observations and includes 13 variables. The variables 'WindDirection', 'Condition' and 'Location' are categorical and the rest are quantitative.

This dataset can be used for various types of weather analysis specific to the cities of Sydney and Canberra during the month of March. Researchers and analysts can examine temperature variations, humidity levels, precipitation patterns, and wind characteristics to understand the unique weather conditions in these two cities during that time frame. It can also be valuable for assessing how different weather conditions may have impacted daily life, outdoor activities, or local businesses in Sydney and Canberra during March. Furthermore, the data can be used to build predictive models for short-term weather forecasting in these specific locations.

----------------------------------
```

```{r}
#----------------------------------#

# 2-  The code for task 2 goes here

summary(weather)


#----------------------------------#

# 3-  The code for task 3 goes here

hist(weather$Temperature)

#----------------------------------#
```
```{asis, echo=TRUE}
4- The reflection and notes for task 2 and 3 goes here

For task 2, the 'summary()' function is really useful in getting an overall overview of the nature of various variable within the dataset we are dealing with. For instance, if we are looking at the Temperature variable, we can observe that the maximum temperature observed is 24.90 degree celcius and the min tempurature recorded is 15 degree celcius, with the mean at 20.05 degree celcius. Also, it is also extremely useful in understanding the categorical variables in the dataset. we have three categorical variables in weather dataset, 'WindDirection', 'Condition' and 'Location'.


For task 3, the histogram of the temperature reading suggests that the distribution of the data is bimodal with one mode between 19 degree celcius and 20 degree celcius and another mode between 23 degree celcius and 24 degree celcius.


----------------------------------

```

## Submission for part B: Vector and Matrix Manipulation
Please follow this structure:

```{r}

library(tidyverse)

#----------------------------------#
# 1-  The code for task 1 goes here
# Create a vector containing the average temperature readings for each month.

# to check the data type of the data column
class(weather$Date)

# Extract the month and year from the "Date" column
weather$Month <- format(weather$Date, "%Y-%m")

# using group_by and summarize to find the monthly average temperature
monthly_average_temp_df <- weather %>%
  group_by(Month) %>%
  summarize(average_temp = mean(Temperature))

print(monthly_average_temp_df)

monthly_average_temperature <- monthly_average_temp_df$average_temp

cat("Q1: Monthly average temperatures: ")
print(monthly_average_temperature)

#----------------------------------#

# 2-  The code for task 2 goes here
# Create two vector containing the average humidity readings for each month; one for Sydney and another for canberra.

# uses group_by and summarize methods to extract the monthly average humidity for canberra snd sydney
monthly_avg_humidity_df <- weather %>%
  group_by(Location, Month) %>%
  summarize(average_humidity = mean(Humidity))

monthly_avg_humidity_df

# weather %>%
#   filter(Month %in% c('2023-04')) %>%
#   select(Month, Location)
# 
# weather %>%
#   select(Date, Month, Location)

canberra_monthly_avg_humidity <- (filter(monthly_avg_humidity_df,Location == 'Canberra'))$average_humidity
# canberra_monthly_avg_humidity <- c(0, canberra_monthly_avg_humidity)

sydney_monthly_average_humidity <- (filter(monthly_avg_humidity_df,Location == 'Sydney'))$average_humidity
# sydney_monthly_average_humidity <- c(sydney_monthly_average_humidity, 0)

cat("Q2: Monthly average humidity for Sydney and Canberra: \n")
print(canberra_monthly_avg_humidity)
print(sydney_monthly_average_humidity)

#----------------------------------#

# 3-  The code for task 3 goes here

# Create a matrix containing the average temperature, humidity, pressure, and wind speed readings for each month.


# first let's calculate the average values of these parameters using tidyverse functions group_by and summarize.
monthly_averages_df <- weather %>%
  group_by(Month) %>%
  summarize(
    average_temp = mean(Temperature), 
    average_humidity = mean(Humidity),
    average_pressure = mean(Pressure),
    average_wind_speed = mean(WindSpeed)
    )

monthly_averages_df <- monthly_averages_df %>% select(-Month)

# as the next step, we will convert the df into a matrix
# reference: https://stat.ethz.ch/R-manual/R-devel/library/base/html/data.matrix.html
monthly_averages <- data.matrix(monthly_averages_df)
# is.matrix(monthly_averages)
monthly_averages




#----------------------------------#

# 4-  The code for task 4 goes here

city_averages_df <- weather %>%
  group_by(Location) %>%
  summarize(
    average_temp = mean(Temperature), 
    average_humidity = mean(Humidity),
    average_pressure = mean(Pressure),
    average_wind_speed = mean(WindSpeed)
    )
city_averages_df

# as the next step, we will convert the df into a matrix
# reference: https://stat.ethz.ch/R-manual/R-devel/library/base/html/data.matrix.html
city_averages <- data.matrix(city_averages_df)
# is.matrix(city_averages)
city_averages



#----------------------------------#

# 5-  The code for task 5 goes here
# Create an array containing the average temperature, humidity, pressure, wind speed, and UV index readings for each day of each month.

weather <- weather %>%
  mutate(DayOfWeek = weekdays(Date))

weather$DayOfWeek

monthly_averages_by_day_df <- 
  weather %>%
  group_by(Month, DayOfWeek) %>%
  summarize(
    average_temp = mean(Temperature), 
    average_humidity = mean(Humidity),
    average_pressure = mean(Pressure),
    average_wind_speed = mean(WindSpeed),
    average_UV_index = mean(UVIndex)
    )

monthly_averages_by_day_df

# create a vector of months
months <- unique(monthly_averages_by_day_df$Month)
months

n_slices <- length(unique(monthly_averages_by_day_df$Month))
n_slices

# calculating col length for the array
n_cols <- ncol(monthly_averages_by_day_df) - 2 # reducing the columns month and weekday
n_cols

# creating a variable to hold row length for the array, which would be no. of days in a week
# length(unique(monthly_averages_by_day_df$DayOfWeek))
n_rows <- 7

# stores days of week in required order
days_of_week <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")

# initializing an empty 3D array
monthly_averages_by_day_arr <- array(NA, c(n_rows, n_cols, n_slices))


# starts the loop
for (i in 1:n_slices) {
  
  # extracting the data for the current month( i.e indexed month in loop) 
  curr_month_df <- monthly_averages_by_day_df %>%
    filter(Month == months[i])
  
  # initialize the data matrix to which average reading for each day will be stored in order
  # the rows of this matrix represents the weekday starting from Monday. 
  # i.e first row represents Monday
  # the columns represents the average readings, it follows the same order as that of the 'monthly_averages_by_day_df' dataframe
  data_matrix <- matrix(data = NA, nrow = 0, ncol = n_cols)
  
  # starts the inner loop, where we will loop through each day of the week
  for( j in  1:n_rows) {
    
    # filtering the current month data frame, based on weekdays so that they can be stored in order to the matrix
    # this will ensure that first row in a slice always represents average readings on mondays of a given month
    curr_day_df <- curr_month_df %>%
      filter(DayOfWeek == days_of_week[j])
    
    # removing month and day columns
    curr_day_df <- curr_day_df[, -which(names(curr_day_df) %in% c("Month", "DayOfWeek"))]
    
    # converting the df row into a vector
    cur_day_vec <- unlist(curr_day_df[1,])
    
    # storing the weekdays(rows) x average readings(cols) to a matrix, where first row represents monday, second tuesday, e.t.c
    data_matrix <- rbind(data_matrix, cur_day_vec)
    
  }
  # print(data_matrix)
  
  # storing the matrix to each slice of the 3D array
  # each slice represents the data of a particular month 
  monthly_averages_by_day_arr[ , ,i] <- data_matrix

}

# prints the 3D array  
print(monthly_averages_by_day_arr)


#----------------------------------#

# 6-  The code for task 6 goes here
# Create an array containing the average temperature, humidity, pressure, wind speed, and UV index readings for each day of each month, for each city.

city_averages_by_day_df <- 
  weather %>%
  group_by(Location, DayOfWeek) %>%
  summarize(
    average_temp = mean(Temperature), 
    average_humidity = mean(Humidity),
    average_pressure = mean(Pressure),
    average_wind_speed = mean(WindSpeed),
    average_UV_index = mean(UVIndex)
    )

city_averages_by_day_df

# create a vector of cities
cities <- unique(city_averages_by_day_df$Location)
cities

n_city_slices <- length(unique(city_averages_by_day_df$Location))
n_city_slices

# calculating col length for the array
n_cols <- ncol(city_averages_by_day_df) - 2 # reducing the columns month and weekday
n_cols

# creating a variable to hold row length for the array, which would be no. of days in a week
# length(unique(monthly_averages_by_day_df$DayOfWeek))
n_rows <- 7

# stores days of week in required order
days_of_week <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")

# initializing an empty 3D array
city_averages_by_day_arr <- array(NA, c(n_rows, n_cols, n_city_slices))


# starts the loop
for (i in 1:n_city_slices) {
  
  # extracting the data for the current city( i.e indexed city in loop) 
  curr_city_df <- city_averages_by_day_df %>%
    filter(Location == cities[i])
  
  # initialize the data matrix to which average reading for each day will be stored in order
  # the rows of this matrix represents the weekday starting from Monday. 
  # i.e first row represents Monday
  # the columns represents the average readings, it follows the same order as that of the 'monthly_averages_by_day_df' dataframe
  data_matrix <- matrix(data = NA, nrow = 0, ncol = n_cols)
  
  # starts the inner loop, where we will loop through each day of the week
  for( j in  1:n_rows) {
    
    # filtering the current month data frame, based on weekdays so that they can be stored in order to the matrix
    # this will ensure that first row in a slice always represents average readings on mondays of a given city
    curr_day_df <- curr_city_df %>%
      filter(DayOfWeek == days_of_week[j])
    
    # removing location/city and day columns
    curr_day_df <- curr_day_df[, -which(names(curr_day_df) %in% c("Location", "DayOfWeek"))]
    
    # converting the df row into a vector
    cur_day_vec <- unlist(curr_day_df[1,])
    
    # storing the weekdays(rows) x average readings(cols) to a matrix, where first row represents monday, second tuesday, e.t.c
    data_matrix <- rbind(data_matrix, cur_day_vec)
    
  }
  # print(data_matrix)
  
  # storing the matrix to each slice of the 3D array
  # each slice represents the data of a particular month 
  city_averages_by_day_arr[ , ,i] <- data_matrix

}

# prints the 3D array  
print(city_averages_by_day_arr)


#----------------------------------#

# 7-  The code for task 7 goes here

# Use matrix multiplication to calculate the product of the transpose of the matrix in task 3 with the vector in task 1.

# vector from task 1
cat("Task 1: vector containing monthly average temperatures: \n")
cat("------------------------------------------------------\n")
print(monthly_average_temperature)
cat("\n")

# matrix from task 3
cat("Task 3: matrix containing the average temperature, humidity, pressure, and wind speed readings for each month: \n")
cat("-------------------------------------------------------------------\n")
print(monthly_averages)
cat("\n")

# matrix transpose
cat("Transpose of the matrix : \n")
cat("------------------------------------------------------\n")
mat_transpose <- t(monthly_averages)
print(mat_transpose)
cat("\n")

# matrix multiplication
product <- mat_transpose %*% monthly_average_temperature
rownames(product) <- NULL
cat("\n Product of matrix multiplication:\n")
cat("---------------------------------\n")
print(product)

#----------------------------------#

```

*** 

## Submission for part C: Looping and Conditional Statements
Please follow this structure:

```{r}
#----------------------------------#

# 1-  The code for task 1 goes here
# Write a loop that calculates the average pressure reading for each month and stores the results in a vector.

# Initialize lists with month names as names
monthly_pressure_readings_sum <- list()
monthly_pressure_readings_count <- list()

for (i in 1:nrow(weather)) {
  key <- weather$Month[i]
  pressure <- as.numeric(weather$Pressure[i])
  
  # Check if the key (month) is already in the lists, and if not, initialize it
  if (!(key %in% names(monthly_pressure_readings_sum))) {
    monthly_pressure_readings_sum[[key]] <- 0
    monthly_pressure_readings_count[[key]] <- 0
  }
  
  # Update the sum and count for the corresponding month
  monthly_pressure_readings_sum[[key]] <- monthly_pressure_readings_sum[[key]] + pressure
  monthly_pressure_readings_count[[key]] <- monthly_pressure_readings_count[[key]] + 1
}

average_monthly_pressure <- c()

for (month in names(monthly_pressure_readings_sum)) {
  average_value <- monthly_pressure_readings_sum[[month]]/monthly_pressure_readings_count[[month]]
  average_monthly_pressure <- c(average_monthly_pressure,average_value)
}

average_monthly_pressure



#----------------------------------#

# 2-  The code for task 2 goes here

# Use a conditional statement to determine how many days had a temperature above 25 degrees Celsius in Sydney.

# variable to store the no. of instances where the Sydney temp is greater than 25 degree celcius
high_temp_count <- 0

# loopint through df
for (i in 1:nrow(weather)) {
  # stores location
  location <- weather$Location[i]
  
  # stores the temperature
  temp <- weather$Temperature[i]
  
  # checks if location is Sydney and temp is greater than 25
  if(location == 'Sydney' & temp > 25) {
    # updates the temp count
    high_temp_count = high_temp_count + 1
  }
}

# print the output
print(high_temp_count)

# to validate
# weather %>%
#   filter(Location == 'Sydney' & Temperature > 25)



#----------------------------------#

# 3-  The code for task 3 goes here
# Use a for loop to calculate the average humidity for the days with a temperature below 21 degrees Celsius in Canberra.

humidity_sum <- 0
days_count <- 0

for (i in 1:nrow(weather)) {
  # stores location
  location <- weather$Location[i]
  
  # stores the temperature
  temp <- weather$Temperature[i]
  
  # checks if location is Sydney and temp is greater than 25
  if(location == 'Canberra' & temp < 21) {
    
    # update days count
    days_count <- days_count + 1
    
    # sum humidity to calculate average
    humidity_sum <- humidity_sum + weather$Humidity[i]
  }
}

average_humidity <- humidity_sum/days_count

average_humidity

# to validate
# weather %>%
#   filter(Location == 'Canberra' & Temperature < 21) %>%
#   summarise(mean_humidity=mean(Humidity))



#----------------------------------#

# 4-  The code for task 4 goes here


# variable to store the high uv days
high_uv_days <- 0

for (i in 1:nrow(weather)) {
  # stores location
  location <- weather$Location[i]
  
  # stores the temperature
  uv_index <- weather$UVIndex[i]
  
  # checks if location is Sydney and temp is greater than 25
  if(location %in% c('Canberra', 'Sydney') & uv_index > 7) {
    
    # update days count
    high_uv_days <- high_uv_days + 1
  }
}

print(high_uv_days)

# to validate
weather %>%
  filter(Location %in% c('Canberra', 'Sydney') & UVIndex > 7) %>%
  summarise(count=n())

#----------------------------------#
```

*** 

## Submission for part D: Data Frame Manipulation
Please follow this structure:

```{r}
#----------------------------------#

# 1-  The code for task 1 goes here



#----------------------------------#

# 2-  The code for task 2 goes here



#----------------------------------#

# 3-  The code for task 3 goes here



#----------------------------------#

# 4-  The code for task 4 goes here



#----------------------------------#
# 5-  The code for task 5 goes here



#----------------------------------#

# 6-  The code for task 6 goes here



#----------------------------------#

# 7-  The code for task 7 goes here



#----------------------------------#

# 8-  The code for task 8 goes here



#----------------------------------#
```

***

## Overall Reflection

Your overall reflection about this assignment goes here ...
