---
title: "Lab 9 - Data Manipulation - dplyr"
author: "Ajul Thomas"
date: "2023-09-26"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}

library(tidyverse)
library(nycflights13)

```



## Excercise 1

```{r}

df <- flights
print(df)

```

```{r}
# Which plane (specified by the tailnum variable) has 
#traveled the most times from New York City airports in 2013?

flights %>% 
  filter(!is.na(air_time)) %>% 
  group_by(tailnum) %>% 
  summarise(count=n()) %>% 
  summarise(
    plain_max = tailnum[which.max(count)], 
    num_times = max(count)
  )


```
```{r}
# What was the average departure delay for the flights per their carriers?

flights %>% 
  filter(!is.na(dep_delay)) %>% 
  group_by(carrier) %>% 
  summarise( count=n(), average_dep_delay=mean(dep_delay))


```

```{r}
flights %>% 
  group_by(dest) %>% 
  summarise(
    max_arr_delay=max(arr_delay, na.rm = TRUE), 
    min_arr_delay=min(arr_delay, na.rm = TRUE), 
    mean_arr_delay=mean(arr_delay, na.rm = TRUE)
    )


```


## Exercise 2

On the flights dataframe, write R code to answer the following questions:


1) What was the average departure delay for each airline?

```{r}


flights %>% 
  group_by(carrier) %>% 
  summarise(avg_dep_delay=mean(dep_delay, na.rm = TRUE)) %>% 
  mutate(name=airlines$name[airlines$carrier==carrier])

```

2) Which airline had the highest average departure delay?

```{r}

flights %>%
  group_by(carrier) %>% 
  summarise(avg_dep_delay=mean(dep_delay, na.rm = T)) %>% 
  summarise(carrier=carrier[which.max(avg_dep_delay)] ,max_avg_dep_delay=max(avg_dep_delay))

```


3) How many flights were delayed in departure or arrival?

```{r}

flights %>% 
  filter(dep_delay > 0 | arr_delay > 0) %>% 
  nrow()
  # summarise(count=n())

```


4) What was the average arrival delay for each destination airport?

```{r}

flights %>% 
  group_by(dest) %>% 
  summarise(mean_arr_delay = mean(arr_delay, na.rm = T))

```

## Exercise 3


We’ll be using the “mpg” dataset, which contains information on fuel economy for various vehicles. Let’s load the necessary libraries for using this dataset

```{r}

library(tidyverse)
data(mpg)

```

Write R chuncks of code to perform the following tasks:
---

1) Let’s start by filtering the data to only include vehicles with a class of “compact”, “subcompact”, or “midsize”, and a transmission type of “automatic”.

```{r}

mpg %>% 
  filter(class %in% c("compact", "subcompact", "midsize") & grepl("auto", trans))

```

2) Let’s group the data by manufacturer and calculate the average city and high way fuel economy for each group.

```{r}

new_mpg_df <- mpg %>%
  filter(class %in% c("compact", "subcompact", "midsize") & grepl("auto", trans)) %>% 
  group_by(manufacturer) %>% 
  summarise(avg_cty=mean(cty, na.rm = T), avg_hwy=mean(hwy, na.rm = T))

```


3) Let’s mutate the data by creating a new column called “avg_combined_mpg” that calculates the average of avg_city_mpg and avg_highway_mpg.

```{r}

combined_mpg_df <- new_mpg_df %>% 
  mutate(avg_combined_mpg = (avg_cty+avg_hwy)/2)

combined_mpg_df

```


4) Let’s view our modified data using the “head” function, sorted by descending avg_combined_mpg.

```{r}

sorted_mpg <- combined_mpg_df %>% 
  arrange(desc(avg_combined_mpg))


head(sorted_mpg)

```


## Exercise 4

On the mpg dataframe, write R code to answer the following:

1) What is the average city and highway fuel economy for each manufacturer?

```{r}

avg_mpg <- mpg %>% 
  group_by(manufacturer) %>% 
  summarise(avg_city=mean(cty, na.rm = T), avg_hwy=mean(hwy, na.rm = T))

```


2) Which manufacturer has the highest overall fuel economy (i.e., the highest average of city and   highway fuel economy)?

```{r}
mpg %>% 
  mutate(overall_mpg=(cty+hwy)/2) %>% 
  group_by(manufacturer) %>% 
  summarise(avg_overall_mpg = mean(overall_mpg)) %>% 
  summarise(
    manufacturer=manufacturer[which.max(avg_overall_mpg)], 
    max_avg_overall_mpg=max(avg_overall_mpg)
    )


```


3) How many cars have a 4-cylinder engine and front-wheel drive?

```{r}

mpg %>% 
  filter(cyl == 4 & drv == 'f') %>% 
  nrow()

```


4) What is the average fuel economy (city and highway) for cars with different engine types and transmission types?

```{r}

mpg %>% 
  group_by(trans, cyl)

```


5) Which car model has the best fuel economy?

```{r}

mpg %>%
  mutate(overall_mpg = (cty+hwy)/2) %>% 
  filter(overall_mpg == max(overall_mpg)) %>% 
  select(manufacturer, model, overall_mpg)


```

