---
title: "Final Assignment, Task 3"
author: "Put your name"
date: "2023-11-2"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Task 3: Data-Driven Modelling: 	(15 marks)

1. Based on the covid19_data dataframe, that you have wrangled and used in the previous tasks, create a separate dataframe named "cor_data" with the data of these variables (CumCases, CumTests, Population, GDP, GDPCapita) variables.

    [Hint: you can use select function on the covid19_data dataframe]

```{r}
# The code and code description of this component go below here

```

2. Compute the correlation matrix between the variables of the “cor_data” and visualise this correlation matrix.

```{r}
# The code and code description of this component go below here

```

3. visualize the distribution of the cumulative cases in the cor_data with and without changing the scale of the x axis to log transformation.

```{r}
# The code and code description of this component go below here

```

4. Divide the cor_data into training and testing, where training data represent 65% of the number of rows.

```{r}
# The code and code description of this component go below here

```

5. Train a linear regression model to predict cumulative cases from the GDP of the countries. Then, evaluate this model on the test data and print the root mean square error value.

```{r}
# The code and code description of this component go below here

```

6. Train another linear regression model to predict cumulative cases from all the other variables. Then, evaluate this model on the test data and print the root mean square error value.

```{r}
# The code and code description of this component go below here

```

7. Interpret the two models and write a small report of highlighting the differences between using the two models. For example, in which cases we should use the first model and in which cases the second one is better to use.

**Interpretation goes below here**:
- 
- 
- 
-

----

**Task 3 final Report**: Highlight the output (Description, graphs and statistics) that have been generated by writing and running the code of the above components. 

----

*** 