---
title: "Final Assignment, Task 3"
author: "Put your name"
date: "2023-11-2"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Task 3: Data-Driven Modelling: 	(15 marks)

1. Based on the covid19_data dataframe, that you have wrangled and used in the previous tasks, create a separate dataframe named "cor_data" with the data of these variables (CumCases, CumTests, Population, GDP, GDPCapita) variables.

    [Hint: you can use select function on the covid19_data dataframe]

```{r}
# The code and code description of this component go below here

cor_data <- covid19_data %>% 
  ungroup() %>% 
  select(CumCases, CumTests, Population, GDP, GDPCapita)
cor_data

```

2. Compute the correlation matrix between the variables of the “cor_data” and visualise this correlation matrix.

```{r}
# The code and code description of this component go below here

cor(cor_data)

ggcorr(cor_data, label = T, label_alpha = T)+
  labs(
    title = "Correlation Matrix"
  )+
  theme(plot.title = element_text(hjust = 0.5)) 

```

3. visualize the distribution of the cumulative cases in the cor_data with and without changing the scale of the x axis to log transformation.

```{r}
# The code and code description of this component go below here

ggplot(cor_data, aes(x=CumCases))+
  geom_histogram(fill = "aquamarine3",color = "#000000")+
  labs(title = "Distribution of Cumulative Cases",
       x = "Cumulative Cases",  # Update x-axis label
       y = "Frequency") +
  theme(plot.title = element_text(hjust = 0.5))

ggplot(cor_data, aes(x=log(CumCases)))+
  geom_histogram(fill = "aquamarine3",color = "#000000")+
  labs(title = "Distribution of Cumulative Cases",
       x = "Logarithm of Cumulative Cases",  # Update x-axis label
       y = "Frequency") +
  theme(plot.title = element_text(hjust = 0.5))

```

4. Divide the cor_data into training and testing, where training data represent 65% of the number of rows.

```{r}
# The code and code description of this component go below here

index <- sample(c(T,F), nrow(cor_data), replace = T, prob = c(.65, .35))

train <- cor_data[index,]
nrow(train)

test <- cor_data[!index,]
nrow(test)

```

5. Train a linear regression model to predict cumulative cases from the GDP of the countries. Then, evaluate this model on the test data and print the root mean square error value.

```{r}
# The code and code description of this component go below here

ggplot(cor_data, aes(x=log(GDP), y=log(CumCases))) +
  geom_point() +
  labs(title = "Distribution of Cumulative Cases vs GDP",
       x = "Logarithm of GDP",  # Update x-axis label
       y = "Logarithm of Cumulative Cases") +
  theme(plot.title = element_text(hjust = 0.5))

model <- lm(CumCases ~ GDP, train)

model

summary(model)

plot(model)

# Testing 

model_predictions <- predict(model, test)

actual_CumCases <- test$CumCases

rmse <- rmse(model_predictions, actual_CumCases)

paste("RMSE of Linear Model 1 :", rmse)

```

6. Train another linear regression model to predict cumulative cases from all the other variables. Then, evaluate this model on the test data and print the root mean square error value.

```{r}
# The code and code description of this component go below here

model2 <- lm(CumCases ~ ., data = train)

summary(model2)

plot(model)

model2_predictions <- 
  predict(model2, test)

rmse_model2 <- rmse(actual_CumCases, model2_predictions)
paste("RMSE of Linear Model 2:", rmse_model2)

```

7. Interpret the two models and write a small report of highlighting the differences between using the two models. For example, in which cases we should use the first model and in which cases the second one is better to use.

**Interpretation goes below here**:
- 
- 
- 
-

1. **Model Complexity**: Model 1 is a simple linear regression with a single predictor (GDP), while Model 2 is a multiple linear regression with multiple predictors (CumTests, Population, GDP, GDPCapita).

2. **Model Performance**: Model 2 (multiple regression) outperforms Model 1 (simple regression) with a higher R-squared value, indicating it explains more variance in CumCases.

3. **RMSE**: Model 2 has a significantly lower RMSE compared to Model 1, suggesting it provides more accurate predictions.

4. **Variable Selection**: Model 1 is appropriate when you have a strong prior belief that a single variable (GDP) drives CumCases. Model 2 is preferred when you suspect that multiple factors (CumTests, Population, GDP, GDPCapita) collectively influence CumCases.

5. **Theoretical Understanding**: Choose Model 1 when you have a well-established theoretical understanding of the relationship between GDP and CumCases. Model 2 is suitable when you believe that other factors also play a role.

6. **Accuracy vs. Simplicity**: Model 2 offers greater accuracy but is more complex. Model 1 is simpler but may sacrifice accuracy.

7. **Research Context**: If precision and predictive power are crucial, Model 2 is the better option.

8. **Data-driven Decision**: If the data supports the inclusion of multiple predictors, Model 2 may be more appropriate. If the data aligns with a single predictor, Model 1 can be effective.

----

**Task 3 final Report**: Highlight the output (Description, graphs and statistics) that have been generated by writing and running the code of the above components. 

----

*** 