---
title: "Lab 3 - Vectors"
author: "Ajul Thomas"
date: "2023-08-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

A vector is a data type for collection of items that are stored together under a name and has a size, which is the number of these items. The c( ) function can be used to create vectors of homogeneous objects by concatenating them together.

To access elements in a vector, you can use the square brackets, i.e [ ] with index/indices to refer to the items you would like to access. The indices can be either positive or negative. Also they can be either scalar or combination of numbers in another vector or sequence.

```{r}

    x <- c(0.5, 0.6)       ## numeric
    print(x)
    
    x <- c(TRUE, FALSE)    ## logical
    print(x)
    
    x <- c(T, F)           ## logical
    print(x)
    
    x <- c("a", "b", "c")  ## character
    print(x)
    
    x <- 9:29              ## integer
    print(x)
    
    x <- c(1+0i, 2+4i)     ## complex
    print(x)



```

## Exercise 2

---
There are built-in vectors in R, such as LETTERS, which is a vector that contains the uppercase letters of the English alphabet.

Add another chunk of code to generate a vector of each of the following:

1) First 12 letters,
2) Odd `numbered’ letters,
3) English vowels (i.e. a, e, i, o, u, y) and
4) English consonants.

```{r}

# built-in functions

letters # lowercase letters
LETTERS # uppercase letters

print(letters[1:12])

```

```{r}

# print odd numbers only
odd_index <- seq(from=1, to=26, by=2)
odd_index

even_index <- seq(from=2, to=26, by=2)
even_index

```

```{r}

letters[odd_index]

vowels <- grep("A|E|I|O|U", LETTERS)
vowels

# printing vowels

letters[vowels]
LETTERS[vowels]

# printing consonants

letters[-vowels]

length(letters[-vowels])

```

## Exercise 3

---

```{r}

weights.people <-  c("Jack",	"Dione", 	"Reda",	"Sally",	"Adam")

weights.before <- c(78,	72,	78,	79,	105)

weights.after <- c(67,	65,	79,	70,	93)

weights.lost <- weights.before - weights.after

weights.lost

weights.lost.mean <- mean(weights.lost)

weights.lost.mean

weights.lost.max.index <- which.max(weights.lost)
weights.lost.max.index

weights.lost.min.index <- which.min(weights.lost)
weights.lost.min.index

paste(weights.people[weights.lost.max.index] , "lost most weight of", max(weights.lost), "Kgs")

paste(weights.people[weights.lost.min.index] , "gained most weight of", min(weights.lost), "Kgs")

```


Alternate method to find max and min index:

```{r}

weights.lost == max(weights.lost)
weights.lost == min(weights.lost)

max_weights_lost_person <-  weights.people[weights.lost == max(weights.lost)]
max_weights_lost_person


min_weights_lost_person <-  weights.people[weights.lost == min(weights.lost)]
min_weights_lost_person

person_weight_gained <-  weights.people[weights.lost < 0]
person_weight_gained


```

```{r}

if(any(weights.lost > 0)) {
  
  print("Someone gained weight")
  
}

```


