---
title: "Lab 11 - Data Visualisation - ggplot2"
author: "Ajul Thomas"
date: "2023-10-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
library(tidyverse)
library(ggplot2)

```



Exercise 1

```{r}

df <- data.frame( gender=factor(rep(c("F", "M"), each=200)),
                 weight=round(c(rnorm(200, mean=55, sd=5), rnorm(200, mean=65, sd=5))))

```


Based on the above code, please perform the following steps:

1)  Create a graph to visualise the spread of the weight values.
```{r}

df <- as_tibble(df)

ggplot(df, aes(x=weight)) + 
  geom_bar(stat = "count")

```
   
2) Create a graph to show how the weight values are spread across the gender groups.
```{r}

weight_spread_gender <- ggplot(df, aes(x=weight, fill=gender)) +
  geom_bar(stat="count", position="dodge")

weight_spread_gender

```
   
3) On the generated graph, can you highlights the average weight per each gender group?
```{r}

means <- df %>% 
  group_by(gender) %>% 
  summarise(average = mean(weight, na.rm = T))

means

female_weight_average <-  means$average[means$gender == 'F']
male_weight_average <-  means$average[means$gender == 'M']

weight_spread_gender +
  geom_vline(aes(xintercept = female_weight_average, linetype="dashed", color="red")) +
  geom_vline(aes(xintercept = male_weight_average, linetype="dashed", color="blue"))

```
    
4) How can you inspect visually whether the weights of each group include any outliers?
```{r}

ggplot(df, aes(x= gender, y=weight)) +
  geom_boxplot(aes(fill=gender))


```

5) what are your insights/findings about the generated analysis?

```{r}



```


Exercise 2

1) Download and import the data called “surveys_complete.csv” from week 11 on Canvas.

```{r}


surveys_df <- read_csv('./data/surveys_complete.csv')
surveys_df

```


2) Generate a graph showing the distribution of the species’ weights based on their id using histogram and boxplot?
```{r}

ggplot(surveys_df)+
  geom_histogram(aes(x=weight, fill=species_id), position = "dodge") +
  labs(x = "Weight", y = " Frequency", title = "Distribution of Species weights by ID")

ggplot(surveys_df, aes(x = species_id, y=weight, fill = species_id)) +
  geom_boxplot() +
  geom_point(alpha = .05, aes(fill = species_id))

ggplot(surveys_df, aes(x = species_id, y=weight, fill = species_id)) +
  geom_boxplot() +
  geom_jitter(alpha = .01, aes(colour = species_id))
```


3) What are your insights based on using the two methods above (ie, boxplots vs histograms)?

```{r}


```




### Exercise 3


Given the following values, (199, 201, 236, 269,271,278,283,291, 301, 303,341):

1)compute the three quartiles and inter-quartile range for this vector.

```{r}

qt_dt <- c(199, 201, 236, 269,271,278,283,291, 301, 303,341)

dt_sum <- summary(qt_dt)

dt_sum

first_qtr <-  dt_sum["1st Qu."]
third_qtr <- dt_sum["3rd Qu."]

iqr <-  third_qtr - first_qtr

data_lowest <- first_qtr - 1.5*iqr
data_highest <- third_qtr - 1.5*iqr

```


2)Identify the outlier values, if any.

```{r}

outliers <- qt_dt[qt_dt < data_lowest | qt_dt > data_highest]

outliers

```


3)Visualise the box-plot as a graph representation for this vector.

```{r}



```


