---
title: "Final Assignment, Task 2"
author: "Put your name"
date: "2023-11-2"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Task 2: Task 2: Exploratory Data Analysis: 		(40 marks)

1. Find the day with the highest death toll reported across the world. Print the date and the death toll of that day. 

```{r}
library(tidyverse)
# The code and code description of this component go below here


highest_death_day <- covid19_data %>%
  group_by(Date) %>%                       # Group by date
  summarise(TotalDeaths = sum(NewDeaths)) %>%   # Sum up deaths for each date
  arrange(-TotalDeaths) %>%                 # Arrange in descending order
  slice(1)                                 # Take the first row

print(highest_death_day)

```


2. Build a graph to show how the cumulative data of (Infected Cases, Deaths, Recovered, Tests) change over the time for the whole world collectively. 

    [Hint: Use geom_line as a geometry function, use log for the Y axis for better presentation, Use different colour to distinguish between new cases, deaths, and recovered]

```{r}
# The code and code description of this component go below here
library(ggplot2)

# Aggregating the data by date
world_data <- covid19_data %>%
  group_by(Date) %>%
  summarise(
    CumCasesWorld = sum(CumCases),
    CumDeathsWorld = sum(CumDeaths),
    CumRecoveredWorld = sum(CumRecovered),
    CumTestsWorld = sum(CumTests)
  )

# Building the plot
ggplot(world_data, aes(x = Date)) +
  geom_line(aes(y = CumCasesWorld, color = "Infected Cases")) +
  geom_line(aes(y = CumDeathsWorld, color = "Deaths")) +
  geom_line(aes(y = CumRecoveredWorld, color = "Recovered")) +
  geom_line(aes(y = CumTestsWorld, color = "Tests")) +
  scale_y_log10() + # Using a logarithmic scale for the Y axis
  labs(
    title = "Cumulative Data Over Time for the Entire World",
    y = "Cumulative Counts (Log Scale)",
    color = "Legend"
  ) +
  theme_minimal()
```


3. Extract the data corresonding to the last day _(05/05/2020)_ for all countries and save it in a separate dataframe and name it _"lastDay_data"_. 

    [Hint: use filter function with Date = "2020-05-05"]

```{r}
# The code and code description of this component go below here

lastDay_data <- covid19_data %>%
  filter(Date == as.Date("2020-05-05"))
lastDay_data
```


4. Based on the data in _"lastDay_data"_ dataframe, extract the corresponding records of the top 10 countries worldwide with current active cases, total confirmed cases, or fatality rate in separate dataframes (_i.e._, top10activeW, top10casesW, top10fatalityW, top10testsMW).

    [Hint: you can use head(arranged_data, n=10) to get the top 10 records and pass the records of  these 10 countries in newly created data frames]

```{r}
# The code and code description of this component go below here
top10activeW <- head(lastDay_data %>% 
                       arrange(desc(Active)),n=10)

top10casesW <- head(lastDay_data %>% 
                      arrange(desc(CumCases)),n=10)

top10fatalityW <- head(lastDay_data %>% 
                         arrange(desc(FatalityRate)),n=10)

top10testMW <- head(lastDay_data %>%
                      arrange(desc(CumTests)),n=10)


```

5. Based on the data of the in _"lastDay_data"_ dataframe, print total confirmed cases, death, recovered cases as well as the total tests per every `continent`. 

```{r}
# The code and code description of this component go below here
continent_totals <- lastDay_data %>%
  group_by(Continent) %>%
  summarise(
    TotalConfirmedCases = sum(CumCases, na.rm = TRUE),
    TotalDeaths = sum(CumDeaths, na.rm = TRUE),
    TotalRecovered = sum(CumRecovered, na.rm = TRUE),
    TotalTests = sum(CumTests, na.rm = TRUE)
  )

print(continent_totals)

```

6. Build a graph to show the total number of cases over the time for the top 10 countries that have been obtained in question 4 (Use log transformation for the values in Y axis for better presentation). 

    [Hint: first you need to get the data of the top-10 countries and then plot their lines, ie, one line per country]

```{r}
# The code and code description of this component go below here
# Get the top 10 country names
top10_countries <- top10casesW$Country

# Filter the main dataframe to include only data for these countries
top10_data_over_time <- covid19_data %>%
  filter(Country %in% top10_countries)

ggplot(top10_data_over_time, aes(x = Date, y = CumCases, color = Country)) +
  geom_line() +
  scale_y_log10() + # Log transformation for Y-axis
  labs(
    title = "Total Number of Cases Over Time for Top 10 Countries",
    y = "Cumulative Cases (Log Scale)",
    x = "Date",
    color = "Country"
  ) +
  theme_minimal()

```

7. Build a graph for the top 10 countries with current highest active cases which was obtained previously in question 4. The graph should have one sub-graph (_i.e._, using facet function) for each of these countries, every sub-graph should show how the new cases, new deaths, and new recovered cases were changing over the time (Use log for Y axis for better presentation, Use different colour to distinguish between new cases, deaths, and recovered). 

    [hint: geom_line function with date on x_axis and each of the values of the variables in y_axis]

```{r}
# The code and code description of this component go below here
# Get the top 10 country names based on active cases
top10_active_countries <- top10activeW$Country

# Filter the main dataframe for these countries
top10_active_data_over_time <- covid19_data %>%
  filter(Country %in% top10_active_countries)


ggplot(top10_active_data_over_time, aes(x = Date)) +
  # Plotting new cases with blue color
  geom_line(aes(y = NewCases, color = "New Cases"), size = 1) +
  # Plotting new deaths with red color
  geom_line(aes(y = NewDeaths, color = "New Deaths"), size = 1) +
  # Plotting new recovered cases with green color
  geom_line(aes(y = Recovered, color = "New Recovered"), size = 1) +
  # Applying a log transformation to the Y-axis for better visualization
  scale_y_log10(
    labels = scales::comma_format(),
    breaks = scales::trans_breaks("log10", function(x) 10^x),
    minor_breaks = scales::trans_breaks("log10", function(x) 10^x)
  ) +
  # Setting a custom color palette
  scale_color_manual(values = c("New Cases" = "blue", "New Deaths" = "red", "New Recovered" = "green")) +
  # Creating a separate sub-graph for each country
  facet_wrap(~ Country, scales = "free_y", ncol = 1) +
  labs(
    title = "New Cases, Deaths, and Recovered Over Time for Top 10 Countries with Highest Active Cases",
    y = "Counts (Log Scale)",
    x = "Date",
    color = "Status"
  ) +
  theme_minimal()

```

8. Build a graph for the top 10 countries with current highest total tests per one million of the population which was obtained previously in question 4. This graph should present total number of infected cases, total tests so far, and the total tests per million of the population for each country. 

    [hint: you can use bar chart to achieve this task]

```{r}
# The code and code description of this component go below here
# Getting the top 10 countries based on tests per million
top10_tests_per_million_countries <- lastDay_data %>%
  arrange(desc(Tests_1M_Pop)) %>%
  head(10) %>%
  pull(Country)

ggplot(top10testMW, aes(x = reorder(Country, Tests_1M_Pop))) +
  # Bar for total infected cases
  geom_bar(aes(y = CumCases, fill = "Total Infected Cases"), stat = "identity", position = "dodge") +
  # Bar for total tests so far
  geom_bar(aes(y = CumTests, fill = "Total Tests"), stat = "identity", position = "dodge") +
  # Text on top of bars to show total tests per million
  geom_text(aes(y = CumTests, label = scales::comma_format()(Tests_1M_Pop)), position = position_dodge(width = 0.9), vjust = -0.25) +
  scale_fill_manual(values = c("Total Infected Cases" = "blue", "Total Tests" = "green")) +
  labs(
    title = "Total Infected Cases, Total Tests, and Tests per Million for Top 10 Countries",
    y = "Counts",
    x = "Country",
    fill = "Metric"
  ) +
  theme_minimal() +
  coord_flip() +
  theme(legend.position = "bottom")



```

9.	Build a graph to present the statistics total, average, median of confirmed cases of the  continents. ( you may use log for Y axis for better presentation, Use Continent in the legend, make sure x-axis labels does not overlap). 

```{r}
# The code and code description of this component go below here

continent_stats <- covid19_data %>%
  group_by(Continent) %>%
  summarise(
    Total = sum(CumCases, na.rm = TRUE),
    Average = mean(CumCases, na.rm = TRUE),
    Median = median(CumCases, na.rm = TRUE)
  )

ggplot(continent_stats, aes(x = reorder(Continent, -Total))) +
  geom_bar(aes(y = Total, fill = "Total"), stat = "identity", position = "dodge") +
  geom_bar(aes(y = Average, fill = "Average"), stat = "identity", position = "dodge") +
  geom_bar(aes(y = Median, fill = "Median"), stat = "identity", position = "dodge") +
  scale_y_log10(
    labels = scales::comma_format(),
    breaks = scales::trans_breaks("log10", function(x) 10^x),
    minor_breaks = scales::trans_breaks("log10", function(x) 10^x)
  ) +
  labs(
    title = "Statistics of Confirmed Cases by Continent",
    y = "Counts (Log Scale)",
    x = "Continent",
    fill = "Statistic"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


```


10. Based on the data of the _"lastDay_data"_ dataframe, list the top 2-countries of each continent that report the highest death toll. 

```{r}
# The code and code description of this component go below here

top2_countries_death <- lastDay_data %>%
  group_by(Continent) %>%
  slice_max(order_by = CumDeaths, n = 2) %>%
  ungroup() %>%
  select(Continent, Country, CumDeaths)

print(top2_countries_death)

```

----

**Task 2 final Report**: For each of the above, ensure that you have extracted the generated statistics (eg, numerical values) or the visualizations (eg, graphs) alongside with your description as output of each component in the knitted report. The marking will be conducted based on the description and validity of these quantitative and qualitative components of the analysis.

----

*** 