---
title: "Assignment 2, Part A"
author: "Put your name"
date: "2023-10-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Part A - Data Importing and Cleaning			(15 marks)

The following code can be used to load the data and merge them all into one data frame, Please run it and then perform the tasks below.

```{r, eval=FALSE}
library(tidyverse)
AU_Date_Format <- "%d/%m/%Y"
US_Date_Format <- "%m/%d/%Y"
files <- list.files("data","*.csv")
act_weather_data <- tibble()

for(i in 1:length(files)){
  data <- read_csv(paste0("data/", files[i]), skip= 7, show_col_types = FALSE)
  assertthat::assert_that(nrow(problems(data)) == 0, 
                  msg="There is still problem/s, which you need to fix first")

  temp <- tryCatch(
            expr    = {  parse_date(data$Date, AU_Date_Format)}, 
            warning = function(e) { parse_date(data$Date, US_Date_Format)})
  

  data$Date <- format(temp, AU_Date_Format)

  act_weather_data <- rbind(act_weather_data, data)
}

# Clean up any temporary variables no longer needed
rm(files, data, temp, i)

```

**Perform the following tasks** 

1. Run the following code to demonstrate the loading of the data has been executed correctly. Provide an explanation in words what each line does (3 marks):
    ```{r eval=FALSE}
dim(act_weather_data)

str(act_weather_data)

act_weather_data %>%
  group_by(Date) %>%
  summarise(count=n()) %>% 
  summarise(max = max(count))

act_weather_data %>%
  summarise_all(funs(sum(is.na(.)))) %>%
  gather() %>%
  filter(value > 0)
```

**Solution**
The above code ...

1. `dim(act_weather_data)`:
   - This line is using the `dim()` function to find the dimensions (number of rows and columns) of the `act_weather_data` dataset. It is checking how many rows and columns are in the data.

2. `str(act_weather_data)`:
   - This line is using the `str()` function to display the structure of the `act_weather_data` dataset. It provides information about the data types and the first few rows of data, giving an overview of what the dataset contains.

3. `act_weather_data %>% group_by(Date) %>% summarise(count = n()) %>% summarise(max = max(count))`:
   - This line is using the pipe operator `%>%` to perform a series of operations on the `act_weather_data` dataset.
   - First, it groups the data by the "Date" column using `group_by(Date)`. This step prepares the data for aggregation by grouping it by unique dates.
   - Then, it calculates the count of rows for each group (in this case, the number of records for each unique date) using `summarise(count = n())`.
   - Finally, it calculates the maximum count (maximum number of records for any date) using `summarise(max = max(count))`.
   - Basically, the above code checks if there are more than one entry against a particular date. As the max count is one, it ensures there is only one entry against each date.

4. `act_weather_data %>% summarise_all(funs(sum(is.na(.)))) %>% gather() %>% filter(value > 0)`:
   - This line also uses the pipe operator `%>%` to perform a sequence of operations on the `act_weather_data` dataset.
   - It starts by using `summarise_all(funs(sum(is.na(.)))` to calculate the sum of missing values (NA) for each column in the dataset.
   - The `gather()` function is used to convert the result into a long format, essentially creating two columns, one for column names and another for the corresponding count of missing values.
   - Finally, it filters the rows where the count of missing values (value column) is greater than 0. This step identifies and shows the columns with missing data.

----

2. Remove the variables, which have no data at all (i.e. all the records in these variables are NAs) (3 marks)

```{r, eval=FALSE}
# solution and documentation of the code goes below this line ...

# returns a logical vector that represents if the varaibales have no data at all
all_cols_na_stats <- colSums(is.na(act_weather_data)) == nrow(act_weather_data)

# extract the names of the variables for which, there is no data at all
na_cols <- names(act_weather_data)[all_cols_na_stats]

# removes the columns for which there is no data at all(i.e. all the records in these variables are NAs)
act_weather_data <- act_weather_data %>% select(-na_cols)


```

----

3. Replace NA values with either the average of its column's values, if they are numeric, or mode if they are categorical. (4 marks)

```{r, eval=FALSE}
# solution and documentation of the code goes below this line ...


# function to calculate mode
mode <-  function(column_name, data = act_weather_data) {
  mode_df <- data %>%
    # removes all the rows na values
    filter(!is.na(!!sym(column_name))) %>%
    # groups by the column of concern
    group_by(!!sym(column_name)) %>%
    # takes the count of unique values
    summarise(n=n()) %>%
    # obtains the values that repeats the most(i.e. mode of the distribution) 
    summarise(mode=(!!sym(column_name))[which.max(n)])
  
  return(mode_df$mode)
}


# get no. of na values in each column
all_cols_na_counts <-  colSums(is.na(act_weather_data))
all_cols_na_counts

# get the names of cols for which there are na values
na_cols_names <- names(all_cols_na_counts[all_cols_na_counts > 0])
na_cols_names

# loops through each column for which there are na values 
for (colmn in na_cols_names) {
  # checks if the column is a categorical variables with mode
  if(!is.numeric(act_weather_data[[colmn]])) {
    act_weather_data <- act_weather_data %>% mutate(!!sym(colmn) := replace_na(!!sym(colmn), mode(colmn)))
  }
  # else it is a numerical column, and replaces na values with the mean
  else {
    act_weather_data <- act_weather_data %>% mutate(!!sym(colmn) := replace_na(!!sym(colmn), mean(!!sym(colmn),na.rm = T)))
  }
}

act_weather_data



```

----
4.	Add two new columns to represent months (`Month`) and years (`Year`) of the data, you may extract the contents of this column from the `Date` column. (2 marks)

```{r, eval=FALSE}
# solution and documentation of the code goes below this line ...

# checks the datatype of the Date column
class(act_weather_data$Date)

# as the type of date column is character, we will format it using as.Date() function 
act_weather_data$Date <-  as.Date(act_weather_data$Date, format=AU_Date_Format)

# check the format of date column again, to ensure everything is as expected
class(act_weather_data$Date)

# seperates the Month and Year column
act_weather_data <- act_weather_data %>% 
  separate(col = Date, into = c('Year', 'Month'), sep = '\\-', remove = F)


```

5. Change the type of the `Month` and `Year` columns to Ordinal with levels as the number of months in a year (i.e. 12) and number of years (i.e., 6). (2 marks)

```{r, eval=FALSE}
# solution and documentation of the code goes below this line ...

# The unique(act_weather_data$Month) function is used to extract the unique values present in the "Month" column.




# converts month column into a factor
act_weather_data$Month <- factor(act_weather_data$Month, levels = sort(unique(act_weather_data$Month)))

# converts year column into a factor
act_weather_data$Year <- factor(act_weather_data$Year, levels = sort(unique(act_weather_data$Year)))

str(act_weather_data)

```

6. Save the weather data to a single file named `act_weather_data_partA.csv`. (1 marks)

```{r, eval=FALSE}
# solution and documentation of the code goes below this line ...

write.csv(act_weather_data, file = "output/act_weather_data_partA.csv")

```

*** 